Respiratory Related Deaths Show Positive Relation to Increasing Air Pollution

### 2024-03-15 paper/paper.qmd
more changes to the paper

---
title: "Respiration Related Death Cause show Signficant Relation to Increasing "
subtitle: "My subtitle if needed"
author: 
  - Vanshika Vanshika
  - Shivank Goel
  - Navya Hooda
thanks: "Code and data are available at: https://github.com/shivankgoel003/Mortality-in-Alberta."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

# Plot the regression graph
ggplot(data, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Graph") +
  theme_minimal()

@@

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

# Plot the regression graph for all lungs causes
ggplot(data_lung, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Lung") +
  theme_minimal()
```

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false
# Plot the regression graph for all heart causes
ggplot(data, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average", y = "Total Deaths", title = "Regression Heart") +
  theme_minimal()

@@

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

# Assuming your data frame is named 'cleaned_avg_PM25_conc' and is already filtered and renamed.
# Convert factors to character to avoid issues in ggplot
data[] <- lapply(data, function(x) if(is.factor(x)) as.character(x) else x)
# Creating the plot with ggplot
p <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = provincial_average, colour = "Provincial Average")) +
  geom_line(aes(y = x10th_percentile, colour = "10th Percentile")) +
  geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
  labs(title = "Annual Trends of PM2.5 Concentrations in Alberta: Provincial Averages and Percentile Ranges (2012-2021)",
       x = "Years", y = "PM2.5 Concentration (μg/m³)") +
  scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
  theme_minimal()

# Display the plot
p

@@

# Plot scatter plot of mortality rates vs. air quality health index
ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~cause) +
  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
       x = "Average Air Quality Health Index",
       y = "Total Deaths")

# Fit the model
model <- glm.nb(total_deaths ~ average_aqi + factor(year) + factor(cause), data = combined_data)

# Model Diagnostics
par(mfrow=c(2,2))
plot(model)

@@

# Plot Observed vs. Predicted Deaths
ggplot(combined_data, aes(x = total_deaths, y = predicted_deaths)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(x = "Observed Deaths", y = "Predicted Deaths") +
  theme_minimal()

# Effect of AQHI on Predicted Deaths
ggplot(combined_data, aes(x = average_aqi, y = predicted_deaths, color = factor(year))) +
  geom_point() +
  geom_line(aes(group = factor(year))) + # Adding lines to connect points by year
  labs(x = "Average Air Quality Health Index", y = "Predicted Deaths") +
  theme_minimal() +
  facet_wrap(~cause)

# Merge datasets on 'year'
combined_data <- merge(deaths_data, average_aqi_by_year, by = "year")

# Plot scatter plot of mortality rates vs. air quality health index
ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~cause) +
  labs(title = "Scatter Plot of Mortality Rates vs. Air Quality Health Index",
       x = "Average Air Quality Health Index",
       y = "Total Deaths")

@@

# Plot for mortality data
ggplot(total_deaths_by_year, aes(x = as.factor(year), y = total_deaths, group = cause, color = cause)) +
  geom_line() +
  labs(title = "Total Deaths by Year for Each Disease",
       x = "Year",
       y = "Total Deaths")

# Ensure the 'year' column is in numeric format for proper line plotting
average_aqi_by_year$year <- as.numeric(as.character(average_aqi_by_year$year))

# Plotting with ggplot2
ggplot(average_aqi_by_year, aes(x = year, y = average_aqi)) +
  geom_line() +  # for lines
  geom_point() + # for points
  labs(title = "Average Air Quality Health Index by Year",
       x = "Year",
       y = "Average Air Quality Index") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks

ggplot(average_aqi_by_year, aes(x = as.factor(year), y = average_aqi, fill = as.factor(year))) +
  geom_bar(stat = "identity") +
  labs(title = "Average Air Quality Health Index by Year",
       x = "Year",
       y = "Average AQHI") +
  theme_minimal()

@@

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Model of "
#| warning: false
library(modelsummary)

modelsummary::modelsummary(
  list(
    "Lung Causes model" = lung_model
  ),
  #statistic = "mad",
  fmt = 2
)

@@

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

@@

pp_check(poisson_model) +
  theme(legend.position = "bottom")

@@

pp_stan_poisson_model <- pp_check(stan_poisson_model)
pp_stan_poisson_model + theme(legend.position = "bottom")

>>>>

---
title: "Respiratory Related Deaths Show Positive Relation to Increasing Air Pollution"
subtitle: "My subtitle if needed"
author: 
  - Vanshika Vanshika
  - Shivank Goel
  - Navya Hooda
thanks: "Code and data are available at: https://github.com/shivankgoel003/Mortality-in-Alberta."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| label: fig-all
#| fig-cap: Provincial Average PM2.5 Quantity Vs Total Deaths
#| echo: false
#| warning: false

# Plot the regression graph
ggplot(data, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average PM2.5", y = "Total Deaths") +
  theme_minimal()

@@

```{r}
#| label: fig-lung
#| fig-cap: Lung Related Causes Mortality Rates vs Average PM2.5 Quantity 
#| echo: false
#| warning: false

# Plot the regression graph for all lungs causes
ggplot(data_lung, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average", y = "Total Deaths") +
  theme_minimal()
```

```{r}
#| label: fig-heart
#| fig-cap: Heart Related Causes Mortality Rates vs Average PM2.5 Quantity 
#| echo: false
#| warning: false
# Plot the regression graph for all heart causes
ggplot(data, aes(x = provincial_average, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line without confidence interval
  labs(x = "Provincial Average", y = "Total Deaths") +
  theme_minimal()

@@

```{r}
#| label: fig-annual
#| fig-cap: Annual Trends of PM2.5 Concentrations in Alberta
#| echo: false
#| warning: false
#| message: false

# Assuming your data frame is named 'cleaned_avg_PM25_conc' and is already filtered and renamed.
# Convert factors to character to avoid issues in ggplot
data[] <- lapply(data, function(x) if(is.factor(x)) as.character(x) else x)
# Creating the plot with ggplot
p <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = provincial_average, colour = "Provincial Average")) +
  geom_line(aes(y = x10th_percentile, colour = "10th Percentile")) +
  geom_line(aes(y = x90th_percentile, colour = "90th Percentile")) +
  labs(x = "Years", y = "PM2.5 Concentration (μg/m³)") +
  scale_colour_manual(values = c("Provincial Average" = "blue", "10th Percentile" = "green", "90th Percentile" = "red")) +
  theme_minimal()

# Display the plot
p
```

```{r}
#| label: fig-four
#| fig-cap: Mortality Rates vs. Air Quality Health Index
#| echo: false
#| warning: false
#| message: false

@@

# Plot scatter plot of mortality rates vs. air quality health index
ggplot(combined_data, aes(x = average_aqi, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~cause) +
  labs(
       x = "Average Air Quality Health Index",
       y = "Total Deaths")
```

```{r}
#| label: fig-fakemodel
#| fig-cap: Model data
#| echo: false
#| warning: false
#| message: false
# Fit the model
model <- glm.nb(total_deaths ~ average_aqi + factor(year) + factor(cause), data = combined_data)

# Model Diagnostics
par(mfrow=c(2,2))
plot(model)
```
```{r}
#| label: fig-fakeprediction
#| fig-cap: Prediction of deaths based on the model
#| echo: false
#| warning: false
#| message: false

@@

# Plot Observed vs. Predicted Deaths
ggplot(combined_data, aes(x = total_deaths, y = predicted_deaths)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(x = "Observed Deaths", y = "Predicted Deaths") +
  theme_minimal()
```

```{r}
#| label: fig-plane
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false
# Effect of AQHI on Predicted Deaths
ggplot(combined_data, aes(x = average_aqi, y = predicted_deaths, color = factor(year))) +
  geom_point() +
  geom_line(aes(group = factor(year))) + # Adding lines to connect points by year
  labs(x = "Average Air Quality Health Index", y = "Predicted Deaths") +
  theme_minimal() +
  facet_wrap(~cause)
```

```{r}
#| label: fig-cause
#| fig-cap: Total Deaths by Year for Each Cause
#| echo: false
#| warning: false
#| message: false

@@

# Plot for mortality data
ggplot(total_deaths_by_year, aes(x = as.factor(year), y = total_deaths, group = cause, color = cause)) +
  geom_line() +
  labs(
       x = "Year",
       y = "Total Deaths") +
  theme(legend.position = "top") +
  theme(legend.text = element_text(size = 5)) +
  guides(color = guide_legend(direction = "vertical"))
```

```{r}
#| label: fig-average_air
#| fig-cap: Average Air Quality Health Index by Year
#| echo: false
#| warning: false
#| message: false
# Ensure the 'year' column is in numeric format for proper line plotting
average_aqi_by_year$year <- as.numeric(as.character(average_aqi_by_year$year))

# Plotting with ggplot2
ggplot(average_aqi_by_year, aes(x = year, y = average_aqi)) +
  geom_line() +  # for lines
  geom_point() + # for points
  labs(
       x = "Year",
       y = "Average Air Quality Index") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) # Optional: Adjust x-axis breaks
```
```{r}
#| label: fig-average_extra
#| fig-cap: Average Air Quality Health Index by Year
#| echo: false
#| warning: false
#| message: false
ggplot(average_aqi_by_year, aes(x = as.factor(year), y = average_aqi, fill = as.factor(year))) +
  geom_bar(stat = "identity") +
  labs(
       x = "Year",
       y = "Average AQHI") +
  theme_minimal()

@@

\begin{align}
y_i | \mu_i, \phi &\sim \text{NegBin}(\mu_i, \phi) \\
\mu_i &= \exp(\alpha + \beta x_i) \\
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta &\sim \text{Normal}(0, 2.5) \\
\phi &\sim \text{Exponential}(1)
\end{align}

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelcauses
#| tbl-cap: "Different causes of mortality and their death counts"
#| warning: false
library(modelsummary)

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelheart
#| tbl-cap: "Heart related causes death count and the air quality"
#| warning: false

```{r}
#| echo: false
#| eval: true
#| label: tbl-modellung
#| tbl-cap: "Lung related causes death count and the air quality"
#| warning: false
modelsummary::modelsummary(
  list(
    "Lung Causes model" = lung_model
  ),
  #statistic = "mad",
  fmt = 2
)

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckfirst
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

@@

pp_check(poisson_model) +
  theme(legend.position = "bottom")
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckneg
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

@@

pp_stan_poisson_model <- pp_check(stan_poisson_model)
pp_stan_poisson_model + theme(legend.position = "bottom")
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckheartlung
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
